---
title: "R4DS Explore II"
author: "Laura Klasek"
date: "7/16/2019"
output: html_document
---
# 4. Workflow: basics
While I read this chapter, I forgot to do the exercises or take notes on it.  

# 5. Data Transformation
```{r package_load}
library(nycflights13)
library(tidyverse)
```

tibble -- data frame tweaked to work better in the tidyverse  

types of variables  
1. int -- integer  
2. dbl -- doubles (real numbers)  
3. chr -- character vectors/strings  
4. dttm -- date-times  
5. lgl -- logical (TRUE or FALSE)  
6. fctr -- factors; categorical variables with fixed possible values  
7. date -- dates  

## 5.2 filter()
Subset observations based on their values  
filter(DATAFRAME, EXPRESSION TO FILTER, ANOTHER EXPRESSION TO FILTER, ETC)  
**to assign to a variable and print the results, wrap assignment in parentheses**  

### 5.2.1 Comparison operators
comparison operators: >, >=, <, <=, ==, !=  
be careful with == and floating point numbers because every number is an approximation: near() instead of ==  

### 5.2.2 Logical operators
& "and"  
| "or"  
! "not"  
xor() "either argument or other argument, not both"  
order of operations is not like English, so be careful  
x %in% y -- selects every row where x is one of the variables in y  

```
nov_dec <- filter(flights, month %in% c(11,12))
```

**__What is De Morgan's law?__**

### 5.2.3 Missing values
missing values (NA) are "contagious" -- any operation with an unknown will also be unknown  
is.na() checks if value is missing  
filter() only includes rows where condition is true and excludes both FALSE and NA  
to preserve missing, ask explicitly  

```
df <- tibble(x = c(1, NA, 3))
filter(df, is.na(x) | x > 1)
```
### 5.2.4 Exercises
1. Find all flights that  
Had an arrival delay of two or more hours
```{r}
#arr_delay is arrival delay in minutes
arr_2h <- filter(flights, arr_delay >= 120)
head(arr_2h)
```
  
Flew to Houston (IAH or HOU)
```{r}
#dest is destination
houston <- filter(flights, dest %in% c("IAH", "HOU"))
head(houston)
```
  
Were operated by United, American, or Delta
```{r}
not_southwest <- filter(flights, carrier %in% c("UA", "AA", "DL"))
head(not_southwest, 10)
```

Departed in summer (July, August, and September)
```{r}
summer <- filter(flights, month == 7 | month == 8 | month == 9)
head(summer)
```

Arrived more than two hours late, but didn’t leave late
```{r}
arrive_late <- filter(flights, arr_delay >= 120 & dep_delay < 0)
head(arrive_late)
```

Were delayed by at least an hour, but made up over 30 minutes in flight
```{r}
delay_catchup <- filter(flights, dep_delay >= 60 & (sched_arr_time - sched_dep_time - 30) > air_time)
```

Departed between midnight and 6am (inclusive)
```{r}
redeye <- filter(flights, dep_time >= 000 & dep_time <= 600)
head(redeye)
```
2. Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?
between() evaluates if values in a numeric vector fall within given range.  It is a shorter way to handle the filter used to generate redeye above.  
```{r}
redeye2 <- filter(flights, between(dep_time, 000, 600))
head(redeye2)
```

3. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?
```{r}
nodep <- filter(flights, is.na(dep_time))
dim(nodep)
View(nodep)
```

8255 flights have a missing dep_time.  They are also missing dep_delay, arr_time, arr_delay, and air_time.  These most likely represent cancelled flights.


## 5.3 arrange()
arrange takes a data frame and set of column names to order by  
if more than one column name is provided, each additional is used to break ties in values of preceding  

desc() reorders a column in descending order  
missing values sorted at the end  

### 5.3.1 Exercises
1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).  
```{r}
missing <- arrange(flights, desc(is.na(dep_time)))
head(missing)
tail(missing)
```

**I have absolutely no idea why this works.**

2. Sort flights to find the most delayed flights. Find the flights that left earliest.
```{r}
most_delayed <- arrange(flights, dep_delay)
View(most_delayed)
```

In this most_delayed tibble, the flights at top are actually the ones that left the earliest (i.e. the ones that have a negative departure delay).  This could be re-sorted with desc(dep_delay) to put the most delayed flights on top.

3. Sort flights to find the fastest flights.
```{r}
speedy <- arrange(flights, air_time)
head(speedy)
```


4. Which flights travelled the longest? Which travelled the shortest?
```{r}
longest <- arrange(flights, desc(distance))
head(longest)

shortest <- arrange(flights, distance)
head(shortest)
```

## 5.4 select()
select() -- focus down on a useful subset of variables  
select(DATAFRAME, name, name, name) selects columns specified by name  
select(DATAFRAME, variable:variable) selects all columns between the two specified  
select(DATAFRAME, -(variable:variable)) selects all columns except those from the two specified  

helpers:  
1. starts_with("abc"): matches names that begin with "abc"  
2. ends_with("xyz"): matches names that end with "xyz"  
3. contains("ijk"): matches names that contain "ijk"  
4. matches("(.)\\1"): selects variables that match a regular expression (ACK, NO, PLEASE NOT THOSE)  
5. num_range("x", 1"3): matches x1, x2, x3  

rename() renames variables and keeps all the variables not explicitly mentioned  
```
rename(flights, tail_num = tailnum)
```

6. everything() -- basically does as on the tin  
```
select(flights, time_hour, air_time, everything())
```

### 5.4.1 Exercises
1. Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.
```{r}
op1 <- select(flights, dep_time, dep_delay, arr_time, arr_delay)
op2 <- select(flights, dep_time:arr_delay)
op3 <- select(flights, starts_with("dep_"), starts_with("arr_"))
op4 <- select(flights, ends_with("_time"), ends_with("_delay"))
op5 <- select(flights, contains("p_t"), contains("d_d"), contains("r_t"), contains("r_d"))
```

2. What happens if you include the name of a variable multiple times in a select() call?
```{r}
multiple_times <- select(flights, dep_time, arr_time, dep_time)
head(multiple_times)
```

It only selects the column once.

3. What does the one_of() function do? Why might it be helpful in conjunction with this vector?
one_of() is a select helper that matches variable names in a character vector.  Basically, you give it a vector and it keeps all of those.  Using it in combintion with the vars allows you to select multiple data sets by the same list of variables

4. Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?
```{r}
select(flights, contains("TIME"))
```

Select helpers ignore case by default.  To change that, specify ignore.case = FALSE in the function.

## 5.5 mutate()
mutate() adds new columns that are functions of existing columns; always adds to end of dataset  

```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)

```

When using mutate, you can refer to the columns you've just created within the same mutate call.  

transmute() does the same thing but keeps only the new variables  

### 5.5.1 Useful create functions
a function used with mutate must be vectorised: takes a vector of values as input, returns a vector with the same number of values as output  
1. arithmetic operators (__remember the recycling rule if vectors are not the same length__); also useful with aggregate functions like sum() or mean()  

2. modular arithmetic: %/% (integer division) and %% remainder -- can be used to break integers into pieces  

3. logs: log(), log2(), log10()  

4. offsets: lead(), lag() refer to the leading or lagging values; compute running differences (x - lag(x)) or find when values change (x != lag(x))  

5. cumulative and rolling aggregates: cumsum(), comprod(), cummin(), cummax(), cummean()  

6. logical comparisons (<, >, etc)  

7. ranking: min_rank() gives smallest first; min_rank(desc()) gives largest first; other options: row_number(), dense_rank(), percent_rank(), cume_dist(), ntile()  

### 5.5.2 Exercises
1. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.
```{r}
dep_reader <- mutate(flights, 
                     dep_time2 = dep_time %/% 100 * 60 + dep_time %% 100)
head(dep_reader)
```

2. Compare air_time with arr_time - dep_time. What do you expect to see? What do you see? What do you need to do to fix it?
```{r}
flights_sml2 <- select(flights, arr_time, dep_time, air_time)

mutate(flights_sml2,
       arr_dep_difference = arr_time - dep_time)
```

You expect the air_time and the difference between the departure and arrival time to be the same.  They are not.  I think this is because all the times are given in local, so we need to somehow take into account the time difference.

3. Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?
```{r}
flights_sml3 <- select(flights, dep_time, sched_dep_time, dep_delay)
head(flights_sml3)
```

The depature delay is the difference between the actual dep_time and the sched_dep_time.  

4. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().  
Documentation: min_rank() uses ties.method = "min" which keeps the ties as ties and sorts them...somehow.  If the second and third ordered values are ties, min_rank() assigns these values to rank 3.
```{r}
super_delay <- mutate(flights,
                    delay_rank = min_rank(dep_delay))
super_delay_sorted <- arrange(super_delay, desc(delay_rank))
```

5. What does 1:3 + 1:10 return? Why?  
1:3 + 1:10 returns c(1+1, 2+2, 3+3, 1+4, 2+5, 3+6, 1+7, 2+8, 3+9, 1+10) because when a vector operation takes place with vectors of unequal length, the shorter vector is recycled.
```{r}
1:3 + 1:10
```

## 5.6 Grouped summaries with summarise()
summarise() collapses a dataframe to a single value based on a calculation you tell it to do  
group_by() -- changes unit of analysis from the complete dataset to individual groups  

### 5.6.1 Combining multiple operations with the pipe
%>% ("then") allows you to chain commands together, avoiding creating new dataframes each time  
```
delays <- flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL")
```

x %>% f(y) turns into f(x, y) and x %>% f(y) %>% g(z) turns into g(f(x, y), z)

### 5.6.2 Missing values
na.rm argument -- if TRUE, removes all the missing values prior to computation by an aggregation function  
If NA included in the aggregation, it multiplies, like syphilus.    

### 5.6.3 Counts
whenever doing any aggregation, include either a count (n()) or a count of non-missing values (sum(!is.na(x)))  

### 5.6.4 Useful summary functions
1. measures of location:  
mean(x) -- sum divided by length  
median(x) -- value where 50% of x is above it  

2. measures of spread:  
sd(x) -- standard deviation  
IQR(x) -- interquartile range  
mad(x) -- median absolute deviation  

3. measures of rank:  
min(x)  
quantile (x, 0.25)  
max(x)  

4. measures of position:  
first(x)  
nth(x, n)  
last(x)  

5. counts
n()  
sum(!is.na(x)) -- number non-missing  
n_distinct() -- number of unique values  
```
not_cancelled %>%
  count(tailnum, wt = distance)
```

6. counts and proportions of logical values
```
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(n_early = sum(dep_time < 500))
```

### 5.6.5 Grouping by multiple variables
group_by(dataframe, GROUP1, GROUP2, GROUP3)  
be careful with progressively rolling up summaries  

### 5.6.6 Ungrouping
ungroup() removes the grouping  

### 5.6.7 Exercises
5. Which carrier has the worst delays?   

```{r}
delays <- flights %>%
  group_by(carrier) %>%
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  arrange(desc(avg_delay))

delays
```

Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights %>% group_by(carrier, dest) %>% summarise(n()))
```{r}
flying_sucks <- flights %>%
  group_by(dest, carrier) %>%
  summarise(
    count = n(),
    avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  arrange(carrier)

flying_sucks
```

```{r}
flying_sucks2 <- flights %>%
  group_by(carrier, dest) %>%
  summarise(
    count = n(),
    avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  arrange(count)

flying_sucks2
```

flying_sucks2 -- not every carrier goes to every destination?

test test hello world why is knitr not working 