---
title: "Swirl Notes"
author: "Laura Klasek"
date: "6/27/2019"
output: html_document
---

To load swirl (must do each time you reload R): library(swirl)
To enter swirl: swirl() 
Follow prompts until swirl asks you to install a course, then select 1 (R Programming)
Choose course 1 (R Programming)
From the lesson menu, work through
1. Basic Building Blocks
(You can Skip 2)
3. Sequences of Numbers
4. Vectors
5. Missing Values
6. Subsetting Vectors
7. Matrices and Data Frames
8. Logic (On this one you can stop if after you get to the 52% completion point)
(You can skip 9-11 for now)
12. Looking at Data

```{r}
library(swirl)
```

```{r}
swirl()
```
You can exit swirl and return to the R prompt (>) at any time by pressing the Esc key. If you are
| already at the prompt, type bye() to exit and save your progress. When you exit properly, you'll see a
| short message letting you know you've done so.

| When you are at the R prompt (>):
| -- Typing skip() allows you to skip the current question.
| -- Typing play() lets you experiment with R on your own; swirl will ignore what you do...
| -- UNTIL you type nxt() which will regain swirl's attention.
| -- Typing bye() causes swirl to exit. Your progress will be saved.
| -- Typing main() returns you to swirl's main menu.
| -- Typing info() displays these options again.


#### 1: Basic Building Blocks
data structure - any object that contains data
vector - simplest type of data structure (this is not a definition)
  create vectors with c() [concatonate or combine]
      uses comma separated lists
      other vectors can be combined
  numeric vectors can be used in arithmetic expressions

arithmetic expressions: +, -, /, *,  ^ (for exponents), sqrt() for square root, and abs() for absolute value
  if using a vector within sqrt(), will act independently on each item in vector
  if two vectors are the same length, R performs arithmetic operation element-by-element
  if vectors are different lengths, R recycles shorter vector
  remember that numeric values are vectors of length 1
  
####2. Workspace and Files
[I know this is optional, but I'm curious.]
getwd() -- get working directory
ls() -- list all objects (i.e. assigned thingamajiggers) in local workspace
list.files() -- list all files in your working directory
dir() -- list all files in your working directory (same as list.files())
args() -- determines the arguments of a function when its called upon that function
dir.create() -- creates directory in the current working directory
    arguments: directory name as string
    to create nested directories, enter file path and set recursive argument to TRUE
setwd() -- set working directory
file.create() -- creates file, filename as string
file.exists() -- checkes to see if file (given as string) exists in the working directory
file.info() -- returns size, whether it is a directory, then more complicated properties
    mode -- file permissions
    mtime -- modification time
    ctime - 'last status change'
    atime - last access times
    exe -- what kind of executable (may be no)
    
    **can use $ to grab specific items** (if know the thing looked for)
    
file.rename(old name, new name) -- renames a file (names given as strings)
file.copy(source, name for copy) -- copies a file
file.path() -- provides relative file path

####3. Sequences of Numbers
n:c -- gives sequence of numbers between n and c, including n and c, increasing in increments of 1
seq() -- gives sequence of numbers
    n, c -- from here to there
    by -- increment
    length -- length of the series to create
length() -- returns length of a thing
to create vector from 1 to N, where N is the length of my_seq
    1:length(my_seq)
    OR
    seq(along.with = my_seq)
    OR
    seq_along()
rep() -- creates vector that replicates first arg as many times as times= arg says
    to repeat elements of a vector passed in, use each= instead of times=

####4. Vectors
atomic vector -- contains exactly one data type
  types: numeric, logical, character, integer, complex
list -- may contain multiple data types

logical operators: <, >, <=, >=, == (exact equality), != (inequality)
A | B -- whether at least one of A or B is true [UNION]
A & B -- whether both true [INTERSECTION]
!A -- negation of A

paste() -- stitches elements of character vector into one continuous string; also stiches elements of multiple character vectors
  collapse = what to put between each element
  sep = what to put between elements of vectors that are being slammed together
  coerces numerics into characters
  
####5. Missing Values
NA = any value that is missing or not available
operations involving NA generally yield NA as result
rnorm(integer) -- returns vector with [integer] draws from standard normal distribution
sample() -- takes sample from given object/vector/c(vectors) of specified size; has argument for replacement
is.na(vector) -- tells whether each element of a vector is NA
be CAUTIOUS when using logical expressions where NAs are around, since NAs can cause havoc there
in R, TRUE = 1 and FALSE = 0, so sum(vector_with_NA) will return count of TRUEs and thus count of NA
NaN -- not a number (i.e. if you divide by zero)
Inf -- infinity

####6. Subsetting Vectors
R uses one-based indexing, not zero-based
"index vector" subset -- use x[a:b] to subset vector x from a index to be index
    logical vectors
      call is.na() or !is.na() on the vector to be subset, which returns a vector of equal length with TRUE for NA (or not NA, depending) -- vector then subsets on indexes that are TRUE
      call a condition (i.e. y > 0) which returns a vector of TRUE/FALSE, subsets on indexes that are TRUE
      if NAs included, includes NAs in resulting vector
    vectors of positive integers
      as noted, R has one-based index
      positive integers return the values with the index corresponding to the integer
    vectors of negative integers
      negative integers return all elements EXCEPT those given as negative
      shorthand: put the '-' in front of the c()
    vectors of character strings
      named elements -- c(foo = 11, bar = 2, norf = NA) -- the names are included
      names(vector) -- returns the names of the elements in the called vector; can also be used to add names by calling and assigning a character of names
      identical() -- tests if two vectors [objects] are identical
      

####7. Matrices and Data Frames
matrices -- contain only single class of data
data frame -- contain single or multiple classes of data

dim() -- returns "dimensions" of object OR sets the dimensions of an R object
  1. to return, call dim(object) 
  2. to set, call dim(object) and assign a vector of dimensions c(rows, columns)
  vectors do not have dimensions -- use length()
  first number = rows, second number = columns
  attributes() will also give you dimensions

class() -- returns the class of object

matrix(data = ?, nrows = , ncols = ) -- makes a matrix

cbind() -- combines columns

data.frame(objects) -- makes a data frame

colnames() -- returns column names OR sets the column names of an R object


####8. Logic
logical expressions can be grouped by parenthesis to specify order of evaluation
`==` equals operator -- tests whether two boolean or number values are equal
`!=` not equals operator -- tests whether two boolean or number values are NOT equal
! NOT operator -- negates logical expressions (i.e. TRUE becomes FALSE)

& (&&) AND operator -- left and right operands TRUE, entire expression is TRUE
  & evalutes across a vector
  && only evaluates first member of vector
| (||) OR operator -- if left or right operand TRUE, entire expression true
  | evaluates across a vector
  || evalutes first member of vector

order of operations
  AND evaluated before OR
  
xor -- exclusive OR

####12. Looking at Data
names(object) -- returns character vector of column names
head(object, #) -- returns specified number of rows (default is 6) from beginning of data frame
tail(object, #) -- returns specified number of rows (default is 6) from end of data frame
summary() -- 
  numeric data -- get minimum, 1st quartile, median, mean, 3rd quartile, maximum
  categorical variable (i.e. factor) -- get number of times each value occurs in the data
table(object$column name) -- return full list of factors with counts
str() -- structure
  class: dimensions
  $each column name : class, preview